---
title: "Data Visualization <br> & R Summary"
subtitle: "Session 5 · MLDS Bootcamp · Fall 2026"
author: "Alice Zhao"
engine: knitr
format:
  revealjs:
    theme: white
    transition: fade
    incremental: true
    loop: true
    echo: true
---

## Discussion

- What have we covered so far during bootcamp?

## Data Visualization Mindset

- Data visualization used to be about syntax
- Now it's about knowing at a high-level:
  1. The core `ggplot` structure
  2. The data visualization workflow
  3. The possibilities
  4. Letting AI fill in the rest

## Agenda

::: nonincremental
- **High-level concepts**
  - Grammar of graphics basics
  - Getting your data ready
  - Categorical data & factors
- **Examples**
  - Basic plot --> AI-assisted tweaks
  - Grouped plot --> AI-assisted tweaks
  - Small multiples
  - Grouped bar chart
:::

## The Grammar of Graphics

- From Leland Wilkinson to Hadley Wickham

::: {.fragment}
```{r}
#| eval: false
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<VARIABLES>))
```
:::

- **data**: a data frame, supplied to `ggplot()`
- **geom**: the shape you want to plot — `geom_point()`, `geom_col()`, `geom_line()`, etc.
- **aes**: the mapping from data to that shape, inside `aes()`
- Layers connect with `+`

## DEMO: A Minimal Plot

```{r}
library(tidyverse)
gapminder <- read_csv("gapminder.csv")
gapminder07 <- gapminder |> filter(year == 2007)
```

::: {.column-body-outset}
```{r}
#| fig-width: 6
#| fig-height: 4
ggplot(gapminder07) +
  geom_point(aes(x = gdpPercap, y = lifeExp))
```
:::

## Getting Your Data Ready

- It's important to mentally lock in what you want your rows and columns to look like **before** plotting
- `ggplot` expects **long data**: one row per observation, one column per variable
  - Remember from `tidyr`: `pivot_longer()` reshapes wide data into long data
  
## EXAMPLE: Getting Your Data Ready

::: {style="font-size: 0.9em"}
```{r}
#| eval: false
# is this data wide or long?
# country      | 1997 | 2002 | 2007
# Afghanistan  | 41.8 | 42.1 | 43.8

# is this data wide or long?
# country      | year | lifeExp
# Afghanistan  | 1997 | 41.8
# Afghanistan  | 2002 | 42.1
```
:::

## EXAMPLE: Getting Your Data Ready

::: {style="font-size: 0.9em"}
```{r}
#| eval: false
# wide -- hard to map to a single geom
# country      | 1997 | 2002 | 2007
# Afghanistan  | 41.8 | 42.1 | 43.8

# long -- one row per country-year, ggplot-ready
# country      | year | lifeExp
# Afghanistan  | 1997 | 41.8
# Afghanistan  | 2002 | 42.1
```
:::

- Most real-world plotting "bugs" are **data-shape** bugs, not **ggplot** bugs
- **Tip**: If your plot looks wrong, check your data's *shape* before you touch the plot code

## EXAMPLE: Getting Your Data Ready

```{r}
head(gapminder, 10)
```

## EXAMPLE: Getting Your Data Ready

```{r}
gapminder |>
  filter(country %in% c("Canada", "Mexico", "United States")) |>
  ggplot() +
  geom_line(aes(x = year, y = lifeExp, color = country))
```

## Categorical Data: Factors

- R stores categorical variables as **factors**
  — a vector with a fixed set of *levels*
- `ggplot` uses factor levels to decide legend order, axis order, etc.
- Without factors, categories are sorted alphabetically — not always what you want

## EXAMPLE: Reordering with Factors

```{r}
gapminder07 |>
  mutate(continent = factor(continent,
    levels = c("Oceania", "Americas", "Europe", "Asia", "Africa"))) |>
  ggplot() +
  geom_boxplot(aes(x = continent, y = lifeExp))
```

## Data Visualization Agenda

::: nonincremental
- **High-level concepts** ✅
  - Grammar of graphics basics ✅
  - Getting your data ready ✅
  - Categorical data & factors ✅
- **Examples**
  - Basic plot --> AI-assisted tweaks
  - Grouped plot --> AI-assisted tweaks
  - Small multiples plot
  - Grouped bar chart plot
:::

## EXAMPLE 1: Create a Basic Plot

- **Goal**: Visualize life expectancy vs GDP
- Your tasks:
  - Sketch what you want your final visual to look like
    - What type of plot?
    - What's on the x and y axis?
  - Figure out what your data needs to look like
    - What should one row represent?
    - What columns do you need?
  
## EXERCISE 1a: Create a Basic Plot

Figure out how to create your plot using `ggplot` syntax

```{r}
#| eval: false
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<VARIABLES>))
```

## SOLUTION 1a: Create a Basic Plot

```{r}
ggplot(gapminder07) +
  geom_point(aes(x = gdpPercap, y = lifeExp))
```

- This is correct, but bare — no title, no labels, default colors

## EXERCISE 1b: Ask AI to Add...

::: nonincremental
- Paste the solution code into your AI assistant of choice
- Ask it to:
  - Add a title and axis labels
  - Color the points by `continent`
  - Apply a clean theme (e.g. `theme_minimal()`)
  - Anything else you'd like
:::

- **Your job**: read the code it gives back — do you understand every line?

## AI SOLUTION: Basic Plot

::: {.column-body-outset}
```{r}
#| fig-width: 8
#| fig-height: 4
ggplot(gapminder07) +
  geom_point(aes(x = gdpPercap, y = lifeExp, col = continent)) +
  labs(
    title = "Life Expectancy vs. GDP per Capita (2007)",
    x = "GDP per capita", y = "Life expectancy"
  ) +
  theme_minimal()
```
:::

## AI SOLUTION: Basic Plot v2

::: {.column-body-outset}
```{r}
#| fig-width: 8
#| fig-height: 4
ggplot(gapminder07) +
  geom_point(aes(x = gdpPercap, y = lifeExp, col = continent), size = 3) +
  labs(
    title = "Wealthier Countries Tend to Live Longer",
    x = "GDP per capita", y = "Life expectancy"
  ) +
  theme_minimal()
```
:::

## EXAMPLE 2: Prep the Data First

- **Goal**: Visualize the total population by continent
- Your tasks:
  - Sketch what you want your final visual to look like
    - What type of plot?
    - What's on the x and y axis?
  - Figure out what your data needs to look like
    - What should one row represent?
    - What columns do you need?

## Data Prep: First Group, Then Plot

- AI can write plotting code with relatively little thought, but you still need to pay close attention to the **data prep** step
- Before you can plot by group, your data has to already be grouped / summarized correctly
- This is where your `dplyr` skills matter most

## EXERCISE 2a: Prep the Data First

::: nonincremental
- Using `gapminder07`, calculate total population (`pop`) by `continent`
:::

## SOLUTION 2a: Prep the Data First

```{r}
pop_by_continent <- gapminder07 |>
  group_by(continent) |>
  summarize(pop = sum(pop))

pop_by_continent
```

## EXERCISE 2b: Create a Grouped Plot

Figure out how to create your plot using `ggplot` syntax

```{r}
#| eval: false
ggplot(data = <DATA>) +
  <GEOM_FUNCTION>(mapping = aes(<VARIABLES>))
```

## SOLUTION 2b: Create a Grouped Plot

```{r}
pop_by_continent |>
  ggplot() +
  geom_col(aes(x = continent, y = pop))
```

## EXERCISE 2c: Ask AI to Add...

::: nonincremental
- Paste the solution code into your AI assistant of choice
- Ask for a bar chart of population by continent sorted highest to lowest, with a Wall Street Journal theme
- Add anything else you'd like
:::

- **Your job**: read the code it gives back — do you understand every line?

## AI SOLUTION: Grouped Plot

```{r}
# install.packages("ggthemes")  # only needed once
library(ggthemes)

pop_by_continent |>
  ggplot() +
  geom_col(aes(x = reorder(continent, -pop), y = pop)) +
  labs(
    title = "Total Population by Continent (2007)",
    x = "Continent", y = "Population"
  ) +
  theme_wsj()
```

## AI SOLUTION: Grouped Plot v2

```{r}
pop_by_continent |>
  ggplot() +
  geom_col(aes(x = pop, y = reorder(continent, pop))) +
  labs(
    title = "Asia Has the Largest Population",
    x = "Population", y = "Continent"
  ) +
  theme_wsj()
```

## Data Visualization Agenda

::: nonincremental
- **High-level concepts** ✅
  - Grammar of graphics basics ✅
  - Getting your data ready ✅
  - Categorical data & factors ✅
- **Examples**
  - Basic plot --> AI-assisted tweaks ✅
  - Grouped plot --> AI-assisted tweaks ✅
  - Small multiples plot
  - Grouped bar chart plot
:::

## BREAK! {.center}

## EXAMPLE 3: Small Multiples

With a partner, try to understand what's happening in this code

```{r}
#| eval: false
ggplot(gapminder07) +
  geom_point(aes(x = gdpPercap, y = lifeExp, size = pop,
                 col = continent), alpha = 0.6) +
  facet_wrap(~continent) +
  labs(
    title = "Life Expectancy vs. GDP per Capita (2007)",
    subtitle = "Point size reflects population",
    x = "GDP per capita", y = "Life expectancy"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

- Once you understand the code, look at the charts and find some insights from the visualization

## EXAMPLE 3: Small Multiples

```{r}
#| echo: false
ggplot(gapminder07) +
  geom_point(aes(x = gdpPercap, y = lifeExp, size = pop,
                 col = continent), alpha = 0.6) +
  facet_wrap(~continent) +
  labs(
    title = "Life Expectancy vs. GDP per Capita (2007)",
    subtitle = "Point size reflects population",
    x = "GDP per capita", y = "Life expectancy"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## What Changed Here?

::: nonincremental
- `size=`  — mapped population to point size
- `alpha=`  — added transparency
- `facet_wrap()` — split into one panel per continent
- `labs()` — added a subtitle
- `theme()` — moved the legend
:::

- None of this is new syntax — it's the same grammar, just more layers

## EXAMPLE 4: Grouped Bar Chart

With a partner, try to understand what's happening in this code

```{r style="font-size:0.9em"}
gapminder07_grouped <- gapminder07 |>
  mutate(life_exp_bracket = case_when(
    lifeExp < 50 ~ "Under 50",
    lifeExp < 60 ~ "50–60",
    lifeExp < 65 ~ "60–65",
    lifeExp < 70 ~ "65–70",
    lifeExp < 75 ~ "70–75",
    TRUE ~ "75+"
  )) |>
  mutate(life_exp_bracket = factor(life_exp_bracket,
    levels = c("Under 50", "50–60", "60–65", "65–70", "70–75", "75+"))) |>
  group_by(continent, life_exp_bracket) |>
  summarize(pop = sum(pop))
head(gapminder07_grouped, 3)
```

## EXAMPLE 4: Grouped Bar Chart

With a partner, try to understand what's happening in this code

```{r style="font-size:0.9em"}
#| eval: false
gapminder07_grouped |>
  ggplot(aes(x = reorder(continent, -pop, sum), y = pop, fill = life_exp_bracket)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Africa Trails, Europe and Oceania Lead in Life Expectancy",
    subtitle = "Population by continent, broken out by life expectancy group (2007)",
    x = NULL, y = "Population", fill = "Life Expectancy"
  ) +
  theme_minimal() +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

## EXAMPLE 4: Grouped Bar Chart

```{r}
#| echo: false
gapminder07_grouped |>
  ggplot(aes(x = reorder(continent, -pop, sum), y = pop, fill = life_exp_bracket)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Africa Trails, Europe and Oceania Lead in Life Expectancy",
    subtitle = "Population by continent, broken out by life expectancy group (2007)",
    x = NULL, y = "Population", fill = "Life Expectancy"
  ) +
  theme_minimal() +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))
```

## What Changed Here?

::: nonincremental
- `fill =` — assigned a bar to each life expectancy group within a continent
- `scales::label_comma()` — adds comma separators to y-axis numbers
- `position = "dodge"` — this is what makes it a **grouped (clustered) bar chart** — bars sit side by side instead of stacking
- `scale_fill_brewer()` — swapped in a color palette
- `theme` — moved location of legend and bolded title
:::

## DISCUSSION

- What are the main steps you need to take when creating a data visualization?
- What are your general thoughts on data visualization?

## Data Visualization Steps

::: nonincremental

- **Sketch the plot**
  - What type of plot
  - What is on the x and y axis
- **Prep the data to create the plot**
  - What should each row look like
  - What should each column look like
- **Create a basic ggplot**
- **Use AI to add to it**
  - Takeaways, themes, etc.
:::

## R Summary {.center}

## R vs RStudio

::: {.fragment .nonincremental}
- **Console, Terminal** — run code
- **Editor** — write scripts and notebooks
- **Environment, History** — current objects, past commands
- **Files, Plots, Packages, Help** — everything else
:::

## R File Types

::: nonincremental
- `.R` — a plain script
- `.Rmd` — code + text *(we've been using a `.qmd` file all along!)*
- `.Rproj` — the project file
- `.csv` — raw / processed data
- `.html` / `.pdf` — rendered output
- `.RDS` — a single saved R object *(vs. `.RData`, which saves your whole environment)*

:::

## R Project Structure

```{.no-scroll}
my-project/
├── my-project.Rproj
├── data/
│   ├── raw/
│   └── processed/
├── R/
│   └── functions.R
├── scripts/
│   ├── 1_clean.R
│   ├── 2_transform.R
│   └── 3_plot.R
├── report.qmd
└── run_all.R
```

## Best Practices

- **Naming:**
  - `snake_case` is the R convention — not camelCase
- **Code Hygiene**:
  - Comment the *why*, not just the *what*
  - Consistent spacing (`x <- 5`, not `x<-5`)
- **Reproducibility:**
  - Restart R often and re-run top to bottom
  - If it doesn't run in a fresh session, it's not done

## And that is your introduction to R! {.center}

## GitHub Follow Ups {.center}

## Last Week

::: nonincremental
- In the GitHub crash course, we covered four workflows:
  - git (**status, add, commit, log**)
  - GitHub: create your own repo (**pull, push**)
  - GitHub: copy an existing repo (**clone, fork**)
  - GitHub: request your changes to get incorporated (**pull request**)
:::

## Q1: Why do we use GitHub?

- **Version control** — a safety net — nothing is ever truly lost
- **Collaboration** — multiple people can edit without overwriting each other
- **Portfolio** — your GitHub is often the first thing an employer looks at

## Q2: What is branching?

- A **fork** is when you copy an existing repo under your own account
  - This is used when you don't have permissions to interact with the original repo
- A **branch** is when you create your own side task *within* a repo
  - You're not editing the main branch directly and can incorporate back in when needed
  - This is used if everyone is closely working on a project
- A **pull request** asks someone to merge your fork or branch back in

## Q3: What happens after I submit a pull request?

- The owner of the original repo can accept or reject your changes
- DEMO: bootcamp-2026 repo

## Q4: What are repo best practices?

- **README** — the front door of a repo, written in *Markdown*
- **Sensible folder structure** — README at the root, data separate from code
- `.gitignore` — set up *before* your first commit to avoid hidden files
- **License** — add a license like MIT for people to use your code
- **Avoid uploading data** — especially large data sets, add links to data or just show results instead of saving in repo

## Q5: What should my repo look like for job applications?

- README does 90% of the work — what, why, key findings, how to reproduce
- Include easily viewable, rendered outputs and graphics, not just source code
- Content ideas:
  - A real question vs a built-in / Kaggle data set
  - A topic you're excited to talk about and break down in interviews

## What's Next

- This afternoon we are going to tie it all together with a bootcamp capstone project
  - R coding: base R and tidyverse
  - R project structure: cleaning, transforming and plotting
  - GitHub pull request

## CLASS ACTIVITY! {.center}

```{=html}
<style>
.small-text {
  font-size: 0.75em;
}

.tight-break {
  line-height: 0.5;
}

.reveal pre {
  margin-top: 10px;
  margin-bottom: 10px;
}

.reveal .column.pad-left {
  padding-left: 50px;
  box-sizing: border-box;
}
</style>
```
