---
title: "New York School Outcomes and County Poverty"
subtitle: "2016 county and school comparisons"
output: html_document
---

## Questions and approach

This report asks how access to free or reduced-price lunch relates to school test performance, and how average test performance differs among counties with low, medium, and high poverty. It also summarizes enrollment and lunch eligibility by county and compares counties with the highest and lowest poverty rates.

The analysis uses 2016, the latest year shared by the school and county poverty data. School test scores are expressed as within-year z-scores because the test scale changes across years. A z-score of zero is the 2016 school average. These comparisons describe associations, not causes.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 5)
library(tidyverse)

school_data <- read_csv(
  "data/processed/final_school_data.csv",
  col_types = cols(school_cd = col_character(), .default = col_guess())
)

data_2016 <- school_data |>
  filter(year == 2016) |>
  mutate(lunch_pct = 100 * (per_free_lunch + per_reduced_lunch))

weighted_avg <- function(x, enrollment) {
  valid <- !is.na(x) & !is.na(enrollment) & enrollment > 0
  if (!any(valid)) return(NA_real_)
  weighted.mean(x[valid], enrollment[valid])
}
```

## County summary

The lunch percentage is weighted by school enrollment. The county poverty rate is taken once per county, since it repeats on every school row. Raw test scores below are enrollment-weighted school averages within 2016.

```{r county-summary}
county_summary = school_data |>
  filter(year == 2016, !is.na(county_name), !is.na(county_per_poverty)) |>
  mutate(per_free_or_reduced = per_free_lunch + per_reduced_lunch) |>
  group_by(county_name) |>
  summarize(
    total_enrollment = sum(total_enroll, na.rm = TRUE),
    pct_free_or_reduced = 100 * weighted_avg(
      per_free_or_reduced, total_enroll
    ),
    pct_poverty = 100 * first(county_per_poverty),
    mean_reading_score = weighted_avg(mean_ela_score, total_enroll),
    mean_math_score = weighted_avg(mean_math_score, total_enroll)
  )

county_summary |>
  select(county_name, total_enrollment, pct_free_or_reduced, pct_poverty) |>
  arrange(desc(pct_poverty))

```

**Takeaway:** The table shows substantial differences in enrollment and poverty across counties. The lunch percentage describes students in the schools in this dataset, and the poverty percentage describes the county population.

### Counties at the poverty extremes

```{r poverty-extremes}
top_bottom_poverty = bind_rows(
  county_summary |>
    slice_max(pct_poverty, n = 5) |>
    mutate(group = "Top 5"),
  county_summary |>
    slice_min(pct_poverty, n = 5) |>
    mutate(group = "Bottom 5")
  ) |>
  select(group, county_name, pct_poverty, pct_free_or_reduced,
         mean_reading_score, mean_math_score)

top_bottom_poverty
```

**Takeaway:** This table makes the county comparison concrete: it places poverty, lunch eligibility, and both test subjects side by side. Raw scores are compared only within 2016; the z-scores below are better for interpreting relative performance.

## Question 1: Lunch eligibility and school test performance

Each point is a school. The lines summarize the overall relationship for each subject.

```{r lunch-performance}
data_2016 = school_data |>
  filter(year == 2016) |>
  mutate(lunch_pct = 100 * (per_free_lunch + per_reduced_lunch))

ggplot(data_2016) +
  geom_point(aes(x = lunch_pct, y = math_z), 
             alpha = 0.8, cex = 0.5) +
  labs(
    title = "Math scores and lunch eligibility in 2016",
    x = "Students eligible for free or reduced lunch (%)",
    y = "Math score (z-score)"
  ) +
  theme_minimal()
```

**Takeaway:** In 2016, schools with a higher percentage of students eligible for free or reduced-price lunch generally had lower math and ELA z-scores. The school-level correlations are about -0.64 for math and -0.61 for ELA. This is an association and does not establish that lunch eligibility causes lower scores.

## Question 2: Test performance by county poverty group

The code first calculates an enrollment-weighted score for each county. It then averages those county scores within each poverty group, giving every county equal weight in the group comparison.

```{r poverty-performance}
group_scores = data_2016 |>
  group_by(county_name, poverty_group) |>
  summarize(math_z = mean(math_z, na.rm = TRUE)) |>
  group_by(poverty_group) |>
  summarize(math_z = mean(math_z, na.rm = TRUE))

ggplot(group_scores) +
  geom_col(aes(x = poverty_group, y = math_z), 
           fill = "steelblue") +
  labs(
    title = "Average math scores by county poverty group",
    x = "County poverty group",
    y = "Average math score (z-score)"
  ) +
  theme_minimal()
```

**Takeaway:** Low-poverty counties have the highest average county scores in both subjects, while high-poverty counties have the lowest. Only three counties fall in the high-poverty group under these cutoffs, so its average should be interpreted cautiously. These patterns do not establish a causal effect of county poverty on test scores.
