---
title: "Poverty and Test Performance in New York Public Schools"
subtitle: "MLDS Boot Camp - Final R Exercise"
author: "Haowen Chen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include = FALSE}
# Global chunk options. This .Rmd sits at the project root, so chunks already
# run from there and the relative paths used in the scripts
# (data/raw/..., data/processed/...) work exactly as they do in a plain
# R session.
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## The question

The Department has asked:

> What can data tell us about the relationship between poverty and test
> performance in New York public schools?

with three follow-ups: how test performance differs between low, medium and
high poverty areas; whether that relationship has changed over time; and
whether access to free / reduced price lunch moderates it.

## Brainstorming (before looking at the data)

**Empathize - why is the stakeholder asking?**
The Department allocates funding and support programs across districts. A
documented poverty-performance gap is the evidence base for where those
dollars go, and the free/reduced price lunch question is really asking
whether an existing intervention appears to help.

**Brainstorm - what would help answer it?**
County-level poverty rates joined to school-level test results; a scatter of
school performance against free/reduced lunch share; average standardized
scores by poverty tier, faceted or coloured by year to show change over time.

**Expected insights.**
I expect a clear negative association between county poverty and test
performance, fairly stable across years, and a lunch-access relationship that
is confounded with poverty rather than obviously protective on its own - the
schools with the most lunch access are also the poorest. Worth stating up
front so I do not read the confound as an effect later.

## Project structure

```
final_r_exercise/
├── final_r_exercise.Rproj
├── run_all.R              # runs the pipeline from a fresh session
├── README.md
├── analysis_notebook.Rmd  # this file
├── data/
│   ├── raw/               # nys_schools.csv, nys_acs.csv (downloaded, untouched)
│   └── processed/         # outputs of the cleaning / merging steps
├── scripts/
│   ├── 0_import.R         # Task 1 - read the raw CSVs
│   ├── 1_clean.R          # Task 3 - missing values, poverty groups, z-scores
│   ├── 2_transform.R      # Task 4 - join schools + ACS
│   ├── 3_analyze.R        # Task 5 - summary tables
│   └── 4_plot.R           # Task 6 - visualizations
└── output/                # saved figures
```

One script per job, each loading the tidyverse itself, so any step can be
re-run without sourcing the whole chain by hand.

## Task 1: Import the data

The two raw files are downloaded into `data/raw/` and left exactly as they
came:

* `nys_schools.csv` - school-level records from the
  [NYS Department of Education](http://data.nysed.gov/downloads.php)
* `nys_acs.csv` - county-level American Community Survey data from the US
  Census Bureau

`read_csv()` is used rather than base `read.csv()`: it returns a tibble, keeps
character columns as characters, and reports the column types it inferred.
The `-99` missing-value code is *not* touched at this stage - the import is a
faithful copy of what is on disk, and recoding belongs in the cleaning script.

```{r import}
source(file.path("scripts", "0_import.R"))
```

## Task 2: Explore the data

```{r explore-schools}
glimpse(nys_schools)
```

`nys_schools` has 35,663 rows and 12 columns: identifiers (`school_cd`,
`school_name`, `district_name`, `county_name`, `region`), `year`, and the
numeric measures (`total_enroll`, `per_free_lunch`, `per_reduced_lunch`,
`per_lep`, `mean_ela_score`, `mean_math_score`). One school-year per row.

```{r explore-acs}
glimpse(nys_acs)
```

`nys_acs` is much smaller - 496 rows, 5 columns - one county-year per row:
`county_name`, `year`, `county_per_poverty`, `median_household_income`,
`county_per_bach`.

**Missing data.** The codebook says missing values are coded `-99` rather
than `NA`, so a plain `is.na()` undercounts them. Counting `-99` directly:

```{r explore-missing}
count_neg99 <- function(x) if (is.numeric(x)) sum(x == -99, na.rm = TRUE) else sum(x == "-99", na.rm = TRUE)

sapply(nys_schools, count_neg99)
```

A few things stood out:

* `mean_ela_score` and `mean_math_score` are `-99` for ~2,200 rows each (about
  6% of school-years) - schools that didn't administer/report a test that
  year.
* Oddly, `-99` also shows up as the *string* `"-99"` in three character
  columns (`county_name`, `district_name`, `region`), all in the same 19
  rows - a school record with no location info at all.
* `district_name` also has real `NA`s (about 1,700, unrelated to `-99`) -
  these turned out to be charter schools, which don't belong to a
  traditional district, so a blank district is legitimate rather than
  missing data.
* `nys_acs` has no `-99` values at all - it's a much cleaner extract.
* `county_per_poverty` ranges from about 4.7% to 30%, with a median around
  12.9%.
* `per_free_lunch` and `per_reduced_lunch` are meant to be proportions
  (0-1), but 44 and 2 rows respectively exceed 1 - most are mild overshoots
  (1.01-1.87, likely rounding/double-counting), plus a couple of extreme
  outliers (257, 53) that are clearly data-entry errors.

```{r explore-pct-over-1}
nys_schools %>% filter(per_free_lunch > 1 | per_reduced_lunch > 1) %>%
  select(school_name, year, per_free_lunch, per_reduced_lunch) %>%
  arrange(desc(per_free_lunch))
```

## Task 3: Recoding and variable manipulation

```{r clean}
source(file.path("scripts", "1_clean.R"))
```

`scripts/1_clean.R` does the following:

**1. Drop exact duplicate rows.** `distinct()` removes 24 fully duplicated
rows in `nys_schools` (identical in every column) before anything else, so
they aren't double-counted later.

**2. Recode `-99` to `NA`.** The three character columns that store `-99`
as a literal string (`county_name`, `district_name`, `region`) are handled
explicitly with `na_if()`; every numeric column is handled in one pass with
`across(where(is.numeric), ~ na_if(.x, -99))`.

**Lunch percentages above 1 set to `NA`.** `per_free_lunch` /
`per_reduced_lunch` should be proportions (0-1); values above 1 (up to 257)
are treated as unreliable and set to `NA` rather than capped, since there's
no way to tell which out-of-range values are close to correct and which are
outright errors.

```{r cap-check}
nys_schools_clean %>% summarise(max_free = max(per_free_lunch, na.rm = TRUE), max_reduced = max(per_reduced_lunch, na.rm = TRUE))
```

**3. Poverty groups.** Counties are split into `"low"` / `"medium"` /
`"high"` poverty tertiles using `dplyr::ntile(county_per_poverty, 3)`,
computed **within each year** (`group_by(year)`). Tertiles were chosen over
a fixed cutoff (e.g. "poverty > 20% = high") because they guarantee
balanced group sizes regardless of how the poverty distribution shifts, and
splitting within year rather than pooling all years keeps "high poverty"
meaning "relatively poor compared to other counties *that year*" - useful
since we also want to look at whether the poverty/performance relationship
holds up over time.

```{r poverty-groups-check}
nys_acs_clean %>% count(year, poverty_group) %>% pivot_wider(names_from = poverty_group, values_from = n)
```

**4. Standardized test scores.** The state's exam changes from year to
year, so raw scale scores aren't comparable across years. `mean_math_score`
and `mean_ela_score` are converted to within-year z-scores
(`group_by(year)` + `scale()`), so a value of e.g. `+1` means "one SD above
that year's average" no matter which year it is.

```{r zscore-check}
nys_schools_clean %>%
  group_by(year) %>%
  summarise(math_mean = mean(math_z, na.rm = TRUE), math_sd = sd(math_z, na.rm = TRUE)) %>%
  head()
```

Each year's `math_z` has mean ~0 and SD 1, as expected.

The cleaned tables are saved to `data/processed/nys_schools_clean.csv` and
`data/processed/nys_acs_clean.csv`.

## Task 4: Merge the datasets

```{r merge}
source(file.path("scripts", "2_transform.R"))
```

**Join key:** `county_name` + `year`. All 62 county names match exactly
between the two files, so no name-cleaning was needed before joining.

**Join type:** `left_join(nys_schools_clean, nys_acs_clean, ...)`, with
schools as the base table. `nys_schools_clean` spans 2008-2017 but
`nys_acs_clean` only covers 2009-2016, so a `left_join` keeps every school
record and leaves the ACS columns `NA` for years outside that range, rather
than an `inner_join` silently dropping those school-years entirely.

```{r merge-check}
nys_merged %>% summarise(
  n_rows          = n(),
  n_no_acs_match  = sum(is.na(county_per_poverty)),
  pct_no_acs_match = round(100 * n_no_acs_match / n_rows, 1)
)
```

About 20% of rows (school-years from 2008 and 2017) have no county-level
data as a result. This is a real coverage gap in the ACS source, not a join
bug -- worth keeping in mind since it means the poverty-related tables and
plots below implicitly cover 2009-2016 only.

The merged dataset is saved to `data/processed/nys_merged.csv`.

## Task 5: Summary tables

```{r analyze}
source(file.path("scripts", "3_analyze.R"))
```

**Table 5.1 -- one row per county:** total enrollment, % of students on
free/reduced lunch, % of the county population in poverty. Each number is a
county-year average first, then averaged again across all years available,
so one county isn't over- or under-represented just because it has more
years of data than another.

```{r table-county-summary}
table_county_summary %>% arrange(desc(pct_poverty))
```

**Table 5.2 -- top 5 / bottom 5 counties by poverty rate:** poverty rate,
free/reduced lunch %, and mean standardized ELA/math performance, side by
side. Z-scores are used because the raw state test scale changes across
years and cannot be averaged meaningfully over the full period.

```{r table-top-bottom-poverty}
table_top_bottom_poverty
```

The two lunch-access columns (`pct_free_reduced_lunch` next to
`pct_poverty`) already hint at the answer to the department's third
question -- they move together almost in lockstep across both groups of
counties.

## Task 6: Data visualization

```{r plots}
source(file.path("scripts", "4_plot.R"))
```

Building these plots surfaced two data-quality issues beyond what Task 3
covered:

* **42 school-years have `per_free_lunch + per_reduced_lunch > 1`** (up to
  186% combined), even though each column individually is <= 1 -- a
  logical impossibility that the single-column check in `1_clean.R` didn't
  catch. These 42 rows are excluded from the scatter plot below only (not
  from the saved processed data), so the x-axis doesn't run past 100%.
* **A handful of schools have extreme z-scores** (as low as -11.7 SD) --
  e.g. EDENWALD SCHOOL, a residential/juvenile-justice facility in the
  Bronx. That's plausibly a genuinely atypical student population rather
  than a data error, so these are left untouched in every table and plot;
  it's noted here as a limitation rather than corrected.

**Plot 1 -- lunch access vs. test performance, at the school level:**

```{r plot1, echo = FALSE}
p1_lunch_vs_performance
```

**Plot 2 -- average test performance by county poverty group:**

```{r plot2, echo = FALSE}
p2_performance_by_poverty
```

**Plot 3 (bonus) -- has the gap changed over time?**

```{r plot3, echo = FALSE}
p3_trend_by_poverty
```

All three figures are also saved to `output/` as PNGs.

## Task 7: Findings

```{r task7-evidence-setup, include = FALSE}
# Recalculate every number used in the conclusions from the analysis data.
# Keeping these values in code prevents the prose from drifting away from
# the figures if the data or earlier cleaning decisions change.
task7_data <- plot_data %>%
  filter(
    !is.na(overall_z),
    !is.na(pct_free_reduced),
    pct_free_reduced <= 1,
    !is.na(county_per_poverty),
    !is.na(poverty_group)
  )

poverty_performance_r <- cor(
  task7_data$county_per_poverty,
  task7_data$overall_z
)
lunch_performance_r <- cor(
  task7_data$pct_free_reduced,
  task7_data$overall_z
)
poverty_lunch_r <- cor(
  task7_data$county_per_poverty,
  task7_data$pct_free_reduced
)

poverty_group_means <- task7_data %>%
  group_by(poverty_group) %>%
  summarise(mean_z = mean(overall_z), .groups = "drop") %>%
  mutate(poverty_group = as.character(poverty_group)) %>%
  tibble::deframe()

low_high_gap <- poverty_group_means[["low"]] -
  poverty_group_means[["high"]]

year_gaps <- task7_data %>%
  group_by(year, poverty_group) %>%
  summarise(mean_z = mean(overall_z), .groups = "drop") %>%
  pivot_wider(names_from = poverty_group, values_from = mean_z) %>%
  mutate(gap = low - high)

gap_2009 <- year_gaps$gap[year_gaps$year == 2009]
gap_2016 <- year_gaps$gap[year_gaps$year == 2016]
gap_reduction <- 1 - gap_2016 / gap_2009

# Exploratory moderation check: both predictors are standardized so the
# interaction is numerically stable. Year indicators absorb broad annual
# shifts. This simple model is descriptive, not a causal estimate.
moderation_data <- task7_data %>%
  mutate(
    poverty_std = as.numeric(scale(county_per_poverty)),
    lunch_std = as.numeric(scale(pct_free_reduced))
  )

moderation_model <- lm(
  overall_z ~ poverty_std * lunch_std + factor(year),
  data = moderation_data
)
moderation_p <- coef(summary(moderation_model))[
  "poverty_std:lunch_std",
  "Pr(>|t|)"
]

task7_evidence <- tibble(
  Measure = c(
    "County poverty vs. test performance",
    "Free/reduced-lunch share vs. test performance",
    "County poverty vs. lunch share",
    "Mean performance: low / medium / high poverty",
    "Low- vs. high-poverty gap: 2009 / 2016",
    "Exploratory poverty x lunch interaction"
  ),
  `Data result` = c(
    sprintf("r = %.2f", poverty_performance_r),
    sprintf("r = %.2f", lunch_performance_r),
    sprintf("r = %.2f", poverty_lunch_r),
    sprintf(
      "%+.2f / %+.2f / %+.2f z",
      poverty_group_means[["low"]],
      poverty_group_means[["medium"]],
      poverty_group_means[["high"]]
    ),
    sprintf("%.2f / %.2f z (%.0f%% narrower)",
            gap_2009, gap_2016, 100 * gap_reduction),
    sprintf("p = %.3f", moderation_p)
  )
)
```

**Conclusion:** Higher-poverty settings have lower average test performance.
The low- versus high-poverty difference is about
`r sprintf("%.2f", low_high_gap)` standard deviations overall, and it remains
visible in 2016 even after narrowing from its 2009 level. Free/reduced-price
lunch eligibility is also strongly associated with lower scores, but this
descriptive analysis does **not** establish that lunch access changes the
strength of the poverty-performance relationship.

### Evidence at a glance

```{r task7-evidence-table, echo = FALSE}
knitr::kable(task7_evidence, align = c("l", "l"))
```

### What can data tell us about the relationship between poverty and test performance in New York public schools?

**Higher poverty is associated with lower test performance, although these
data do not prove that poverty causes lower scores.** Across the complete
school-year records, county poverty and average math/ELA performance have a
correlation of **r = `r sprintf("%.2f", poverty_performance_r)`**. At the
school level, the share of students qualifying for free/reduced-price lunch
has an even stronger negative correlation with performance
(**r = `r sprintf("%.2f", lunch_performance_r)`**; Plot 1). The spread of
scores at every poverty and lunch level shows that economic disadvantage is
important, but it is not the only factor related to school performance.

### What's the difference in test performance between low, medium and high poverty areas?

**Performance decreases at each step from low- to high-poverty counties.**
Average school-level math/ELA performance is
**`r sprintf("%+.2f", poverty_group_means[["low"]])` z** in low-poverty
counties, **`r sprintf("%+.2f", poverty_group_means[["medium"]])` z** in
medium-poverty counties, and
**`r sprintf("%+.2f", poverty_group_means[["high"]])` z** in high-poverty
counties (Plot 2). The low- versus high-poverty difference is therefore
**`r sprintf("%.2f", low_high_gap)` standard deviations**, a substantial
descriptive gap.

### Has this relationship changed over time?

**The gap is smaller at the end of the period, but it does not disappear.**
The low- versus high-poverty difference falls from
**`r sprintf("%.2f", gap_2009)` z in 2009** to
**`r sprintf("%.2f", gap_2016)` z in 2016**, a reduction of approximately
**`r sprintf("%.0f", 100 * gap_reduction)`%** (Plot 3). This is an endpoint
comparison rather than evidence of a steady or causal trend, but it shows
that the observed gap was meaningfully narrower in 2016 while remaining
large.

### Is this relationship at all moderated by access to free/reduced price lunch?

**This analysis does not find clear evidence that lunch eligibility
moderates the poverty-performance relationship.** County poverty and the
school free/reduced-lunch share are themselves strongly related
(**r = `r sprintf("%.2f", poverty_lunch_r)`**), which makes their separate
roles difficult to isolate. In an exploratory model with poverty, lunch
share, their interaction, and year indicators, the interaction has
**p = `r sprintf("%.3f", moderation_p)`**. That is just above the conventional
0.05 threshold, so the correct conclusion is **inconclusive evidence of
moderation**, not proof that moderation is absent. A stronger answer would
require a model designed for repeated school observations and additional
school/community covariates.

### Limitations

* The merged dataset only has complete poverty/lunch data for 2009-2016;
  2008 and 2017 school records exist but can't be placed into a poverty
  group.
* A few schools (e.g. residential/juvenile-justice facilities) have
  extreme scores that weren't corrected or removed -- they're real data
  points, but atypical ones, and readers should keep that in mind since
  they are included in the analysis above.
* Everything here is a correlational read of aggregated school- and
  county-level averages, not a causal or student-level analysis -- it says
  where the gaps are, not what's driving them or what would close them.
* The moderation model is exploratory and treats repeated school-year
  observations as independent. Its p-value should be read as a screening
  result, not a definitive policy finding.
