---
title: "NYS Schools and County Poverty Analysis"
output: html_document
---

# Introduction

This analysis examines the relationship between socioeconomic conditions
and student academic performance in New York State. I use school-level
data from the New York State Department of Education and county-level
data from the American Community Survey (ACS).

The analysis focuses on county poverty, students' access to free or
reduced-price lunch, and standardized math and English Language Arts
(ELA) performance.

# Data Preparation

The raw datasets were first cleaned before analysis.

Missing values were originally coded as `-99`, so these values were
converted to `NA`. I also created a categorical poverty variable that
classifies counties as low, medium, or high poverty.

Because the NYS tests changed over time, raw test scores are not directly
comparable across years. Therefore, math and ELA scores were standardized
within each year using z-scores.

The school data and ACS data were then merged using county name and year.

# Summary Tables

The following tables summarize socioeconomic conditions and academic
performance across New York counties.

```{r summary-tables, message=FALSE, warning=FALSE}
# Run the analysis script that creates the summary tables
source("scripts/analyze.R")
```

## County Summary

This table shows total enrollment, the average percentage of students
receiving free or reduced-price lunch, and the average county poverty rate.

```{r county-summary}
county_summary
```

## Counties with the Highest Poverty Rates

```{r top-poverty}
top5_table
```

## Counties with the Lowest Poverty Rates

```{r bottom-poverty}
bottom5_table
```

# Data Visualizations

The following figures explore the relationship between socioeconomic
conditions and student performance.

```{r load-plots, message=FALSE, warning=FALSE}
# Run the plotting script
source("scripts/plot.R")
```

# Findings

Based on the summary tables and visualizations, there appears to be a
relationship between socioeconomic conditions and academic performance.

Counties with higher poverty rates tend to have a larger proportion of
students receiving free or reduced-price lunch.

The figures also allow us to compare standardized math and ELA performance
across low-, medium-, and high-poverty counties.

Because test scores were standardized within each year, the z-scores
represent a school's performance relative to other schools in the same year.

# Conclusion

Overall, the analysis demonstrates how school-level educational outcomes
can be combined with county-level socioeconomic information to examine
patterns in student performance. The use of standardized test scores makes
comparisons across years more meaningful.