---
title: "Poverty and Test Performance in New York Public Schools"
author: "Lilibet Trautman"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

## Question

**What can data tell us about the relationship between poverty and test performance
in New York public schools?**

I combined two datasets: school-level test scores and lunch data from the NYS
Department of Education, and county-level poverty data from the US Census (ACS).
Because the state test changes over time, all scores are standardized to a
**z-score within each year**, so a value of 0 means "average for that year" and
+1 means "one standard deviation above average."

```{r load-data, echo=FALSE}
merged <- read_csv("data/processed/schools_acs_merged.csv")
```

## Question 1: Do low, medium, and high poverty areas differ in test performance?

Counties were split into equal thirds (tertiles) by their poverty rate, so each
group holds roughly the same number of counties.

```{r q1-table, echo=FALSE}
merged |>
  filter(!is.na(poverty_group)) |>
  group_by(poverty_group) |>
  summarize(
    n_schools     = n(),
    avg_z_math    = mean(z_math, na.rm = TRUE),
    avg_z_ela     = mean(z_ela,  na.rm = TRUE),
    avg_free_red  = mean(per_free_lunch + per_reduced_lunch, na.rm = TRUE)
  )
```

```{r q1-plot, echo=FALSE, fig.width=7, fig.height=5}
merged |>
  filter(!is.na(poverty_group)) |>
  group_by(poverty_group) |>
  summarize(avg_z_math = mean(z_math, na.rm = TRUE)) |>
  ggplot(aes(x = poverty_group, y = avg_z_math)) +
  geom_col(fill = "steelblue") +
  labs(
    x = "County poverty group",
    y = "Average math score (z-score)",
    title = "Higher-Poverty Counties Have Lower Average Test Scores"
  ) +
  theme_minimal()
```

**Takeaway:** Test performance falls as county poverty rises. High-poverty
counties score below the state average; low-poverty counties score above it.

## Question 2: Has this relationship changed over time?

```{r q2-plot, echo=FALSE, fig.width=7, fig.height=5}
merged |>
  filter(!is.na(poverty_group)) |>
  group_by(year, poverty_group) |>
  summarize(avg_z_math = mean(z_math, na.rm = TRUE), .groups = "drop") |>
  ggplot(aes(x = year, y = avg_z_math, color = poverty_group)) +
  geom_line(linewidth = 1) +
  labs(
    x = "Year",
    y = "Average math score (z-score)",
    color = "Poverty group",
    title = "The Poverty Gap in Test Scores Persists Across Years"
  ) +
  theme_minimal()
```

**Takeaway:** The gap between low- and high-poverty counties is stable over the
period, not closing. Poverty tracks with lower scores year after year.

## Question 3: Is the relationship moderated by access to free/reduced lunch?

At the school level, the share of students on free/reduced lunch (a direct
poverty measure) has a clear negative relationship with test scores.

```{r q3-plot, echo=FALSE, fig.width=7, fig.height=5}
merged |>
  mutate(per_free_reduced = per_free_lunch + per_reduced_lunch) |>
  ggplot(aes(x = per_free_reduced, y = z_math)) +
  geom_point(alpha = 0.05, color = "steelblue") +
  geom_smooth(method = "lm", color = "darkred") +
  labs(
    x = "Share of students on free/reduced lunch",
    y = "Math score (z-score within year)",
    title = "Schools with More Free/Reduced Lunch Score Lower in Math"
  ) +
  theme_minimal()
```

```{r q3-corr, echo=FALSE}
merged |>
  mutate(per_free_reduced = per_free_lunch + per_reduced_lunch) |>
  summarize(correlation = cor(per_free_reduced, z_math, use = "complete.obs"))
```

**Takeaway:** Free/reduced lunch is itself a strong marker of disadvantage and
moves with lower scores. Rather than softening the poverty effect, it reinforces
the same story from the school side that the county poverty data shows from the
area side.

## Caveats

- ACS county data only covers 2009-2016, so schools in 2008 and 2017 have no
  poverty match and are dropped from the poverty-group analyses.
- This is correlation, not causation -- poverty does not directly *cause* lower
  scores; it is bound up with many other factors (funding, resources, etc.).
