---
title: "Base R"
subtitle: "Session 2 · MLDS Bootcamp · Fall 2026"
author: "Alice Zhao"
engine: knitr
format:
  revealjs:
    theme: white
    transition: fade
    incremental: true
    loop: true
    echo: true
---

## Reshuffle #4

- Find a new group of 2-3 people
- Introduce yourselves (again!)
- Quickly recap what we discussed this morning

## Session 1 Recap

::: nonincremental
- Math, variables, functions
- Data types: numeric, character, logical
- Packages
- Data structures: vectors, lists, factors, data frames
- Reading in data
:::

## Session 2 Mindset

- We covered foundational concepts in Session 1
  - Working with the RStudio UI
  - Loading data into data frames
- Now we are moving to **base R**: R code without any additional packages
- Why learn base R?
  - This helps you understand how R works
  - Many packages rely on base R code
  - Useful for simple tasks

## Session 2 Setup

- Clean up your console and environment before we start
  - Click the broom icons in both tabs
- We'll be switching from R scripts to R notebooks
  - `R script (.R)`: plain code you can use as a scratch pad
  - `R markdown (.Rmd)`: code along with text documentation, that you "knit" together
  - `R notebook (.Rmd)`: same as markdown, but can run "chunks" of code as you go *(great for data analysis!)*

## Agenda

::: nonincremental
- Reading in data
- Data wrangling
- Statistics
- Data visualization
- Writing files
:::

## EXERCISE: Reading in Data

- Read in the `gapminder.csv` file from this morning using base R only

::: {.fragment}
```{r}
setwd('~/Desktop')
df <- read.csv('gapminder.csv')
head(df)
```
:::

## EXERCISE: Get a Feel for Your Data

::: nonincremental
- Before doing anything with a new data set, get a feel for it
- Useful functions:
  - `str()` — high-level structure
  - `dim()` — number of rows and columns
  - `names()` — column names
  - `head()` / `tail()` — preview the data
  - `View()` — preview the data in the UI
:::

## SOLUTION: Get a Feel for Your Data

```{r style="font-size: 0.9em;"}
str(df)
dim(df)
names(df)
head(df, 1)
```

## Exploring Data Frames

- `summary()` gives summary statistics for every column
- `mean()`, `median()`, `var()`, `sd()`, `quantile()` — useful on individual columns
- `table()` builds a frequency table for categorical columns
- `prop.table()` turns a frequency table into proportions

## EXERCISE: Exploring Data Frames

::: nonincremental
- Interpret the summary statistics for all columns
- Find the average population
- Run `table(df$year)` -- how would you interpret the output?
- Run `prop.table(table(df$continent))` -- how would you interpret the output?
:::

## SOLUTION: Exploring Data Frames

```{r}
summary(df)
mean(df$pop)
```

## SOLUTION: Exploring Data Frames

```{r}
table(df$year)
prop.table(table(df$continent))
```

## Subsetting

- Subsetting means working with just part of a data frame
- Two common approaches:
  - Index it directly: `df[df$continent == "Asia", ]`
  - Use `subset()`: `subset(df, subset = continent == "Asia")`

## Sorting

- `sort()` reorders values
  - Ascending by default
  - Use `decreasing = TRUE` to flip
- `order()` gives you the index positions in sorted order
  - Useful for reordering a whole data frame
- **Rule of thumb**
  - Use `sort()` for vectors and `order()` for data frames

## EXERCISE: Subsetting & Sorting

::: nonincremental
- Make a new data frame called `df07` with just the 2007 rows
- Create a table that shows how many countries are in each continent
- Sort the continents by how many countries are in each
- Look up Mexico's population in 2007
:::

## SOLUTION: Subsetting & Sorting

```{r style="font-size: 0.8em;"}
df07 <- df[df$year == 2007, ]
head(df07, 3)
table(df07$continent)
sort(table(df07$continent))
df07$pop[df07$country == "Mexico"]
```

## EXERCISE: Bonus

::: nonincremental
- Sort the 2007 data by population descending
:::

## SOLUTION: Bonus

```{r}
df07$rownum <- 1:nrow(df07)
head(df07)
order(df07$pop)
```

## SOLUTION: Bonus

```{r}
head(df07[order(df07$pop), ])
head(df07[order(df07$pop, decreasing = TRUE), ])
```

## Reshuffle #5

::: nonincremental
- Find a new group of 2-3 people
- Introduce yourselves
:::

## Adding & Removing Columns

- Add a new column two ways:

::: {.fragment}
```{r}
#| eval: false
df$newvar <- newvar
df <- cbind(df, newvar)
```
:::

- Remove a column just as easily:

::: {.fragment}
```{r}
#| eval: false
df$newvar <- NULL
```
:::

## Creating New Columns

- Two common approaches:
  - Simple case: one function does the job
     - Just assign its result to a new column
  - Conditional case: no single function fits
     - Create an empty column, and then fill it in based on a condition

## EXERCISE: Creating New Columns

::: nonincremental
- Round `lifeExp` using `round()` and store it as `lifeExp_round`
- Create `lifeExp_highlow`
  - Start it as empty (all `NA` values)
  - Then fill in `High` where life expectancy is above the mean and `Low` where it's below
  - Check your work with `table()` to see how many High and Low values there are
:::

## SOLUTION: Creating New Columns

```{r}
df07$lifeExp_round <- round(df07$lifeExp)
head(df07)
df07$lifeExp_highlow <- NA
df07$lifeExp_highlow[df07$lifeExp > mean(df07$lifeExp)] <- "High"
df07$lifeExp_highlow[df07$lifeExp < mean(df07$lifeExp)] <- "Low"
table(df07$lifeExp_highlow)
```

## Aggregating

- We often want summary statistics *by group*
- `aggregate()` does this: `aggregate(y ~ x, data = df, FUN = mean)` gives the mean of `y` for each group in `x`

## EXERCISE: Aggregating

::: nonincremental
- Find the mean life expectancy in 2007 for each continent
:::

## SOLUTION: Aggregating

```{r}
aggregate(lifeExp ~ continent, data = df07, FUN = mean)
```

## EXERCISE: More Function Practice

::: nonincremental
- Create a new column called `decade`
- Display the unique decades in the data
- How many unique decades are there?
- Sort the unique decades
:::

## SOLUTION: More Function Practice

```{r style="font-size: 0.9em;"}
df$decade <- round(df$year, -1)
head(df)
unique(df$decade)
length(unique(df$decade))
sort(unique(df$decade))
```

## Data Wrangling Review

:::: {columns}

::: {.column width="60%" .fragment .nonincremental}
- Read in data
- Get a feel for your data
- Explore data
- Subset data
- Sort data
- Add, remove and create new columns
- Aggregate data
- More functions: unique, length, etc.
:::

::: {.column width="40%" .small-text .pad-left .fragment .nonincremental}
> Shortcuts
> 
> - cmd/ctrl + return
> - up/down
> - cmd/ctrl + left/right
> - tab
:::

:::::

## BREAK! {.center}

## Reshuffle #6

::: nonincremental
- Find a new group of 2-3 people
- Introduce yourselves
:::

## Agenda

::: nonincremental
- Reading in data ✅
- Data wrangling ✅
- Statistics
- Data visualization
- Writing files
:::

## Question

- What's the relationship between a country's average **life expectancy** vs their **GDP** and **population**?

::: {.fragment}
```{r}
df07
```
:::

## Brainstorm

- How would you go about answering this question?

<br>

- **Statistics!**
  - Correlations and modeling are built into base R

## Correlation

- `cor()` measures the linear relationship between two numeric variables (between -1 and 1)
- `plot()` shows you the visual representation of the data

<br>

- **EXERCISE**:
  - Use `cor()` and `plot()` to compare the relationships between `life expectancy`, `GDP` and `population`
  - What conclusions can you make?

## Correlation: Life Exp vs GDP

```{r}
cor(df07$lifeExp, df07$gdpPercap)
```

```{r}
plot(df07$lifeExp, df07$gdpPercap)
```

## Correlation: Life Exp vs Population

```{r}
cor(df07$lifeExp, df07$pop)
```

```{r}
plot(df07$lifeExp, df07$pop)
```

## Correlation: Life Exp vs Population v2
```{r style="font-size: 0.8em;"}
df07[df07$pop > 1000000000, c("country", "pop")]
plot(df07$lifeExp[df07$pop < 1000000000], df07$pop[df07$pop < 1000000000])
```

## Linear Regression

- `lm()` fits a linear regression model
- Formula syntax: `lm(y ~ x1 + x2, data = df)`
- `summary()` gives coefficients, p-values, and model fit

## Linear Regression

::: {.fragment style="font-size: 0.95em;"}
```{r}
reg <- lm(lifeExp ~ gdpPercap + pop, data = df07)
summary(reg)
```
:::

## Reshuffle #7

::: nonincremental
- Find a new group of 2-3 people
- Introduce yourselves
:::

## Data Visualization

- We'll cover visualization briefly today using **base R**
  - We'll go deeper with `ggplot2` in a later session
- Why start with base R?
  - Great for quickly exploring data while you're working
  - Its syntax shows up again in other packages
  
## EXERCISE: Data Visualization

- One person will do the histogram exercise
- One person will do the scatter plot exercise
- Then teach the other person what you learned

## EXERCISE: Histograms

::: nonincremental
- Create a histogram of the variable `lifeExp` in `gapminder07`
- What does the `breaks` parameter do? Try modifying it.
::: 

```{r}
hist(df07$lifeExp,
     main = "Distribution of life expectancy across countries in 2007",
     xlab = "Life expectancy", ylab = "Frequency", breaks=10)
```

## Scatter Plots

::: nonincremental
- Create a scatter plot of `lifeExp` vs `gdpPercap` using the `gapminder07` data
- What is the `abline`? Try modifying it.
:::

```{r}
plot(lifeExp ~ gdpPercap, data = df07,
     main = "Relationship between life expectancy and GDP per capita in 2007",
     ylab = "Life expectancy", xlab = "GDP per capita")
abline(h = mean(df07$lifeExp))
```

## Writing to Files

- Save a cleaned-up data frame with `write.csv()`
  - *remember `read.csv()` from earlier*

::: {.fragment}
```{r}
#| eval: false
write_csv(df07, file = ".csv")
```
:::

<br> 

- You can **save your environment objects** as an `.RData` file by clicking the `Save` icon in the Environment tab

## Discussion

- Combine with another local group
  - What did we cover today?
  - Anything stick out in particular?

<br>
  
::: {.fragment}
**CLASS ACTIVITY #2!**
:::

```{=html}
<style>
.small-text {
  font-size: 0.75em;
}

.tight-break {
  line-height: 0.5;
}

.reveal pre {
  margin-top: 10px;
  margin-bottom: 10px;
}

.reveal .column.pad-left {
  padding-left: 50px;
  box-sizing: border-box;
}
</style>
```
