---
title: "R Foundations"
subtitle: "Session 1 · MLDS Bootcamp · Fall 2026"
author: "Alice Zhao"
engine: knitr
format:
  revealjs:
    theme: white
    transition: fade
    incremental: true
    loop: true
    echo: true
---

## Why R?

- R is open source
- R has extensive statistical capabilities
- R can create custom, publication-quality visualizations
- R has a large, active community of users

## Discussion

- Who has used R before?
- What are your thoughts on R?

## What is RStudio?

- It's an integrated development environment (IDE) for R
- It provides a user-friendly interface to write and run code
- RStudio Desktop is the most common way people use R\
  (vs RStudio Server or Posit Cloud)
- Posit (formerly RStudio) is the company that maintains it

## Demo

::::::: columns
::: {.column width="50%"}
- R in the Terminal
:::

::::: {.column width="50%"}
- RStudio IDE

:::: {.fragment .nonincremental .small-text}
::: {}
> - Console
>
> - Editor
>
>   - Create a new R Script
>   - Run vs Source
>
> - Environment & History
>
> - Files
>
>   - Set Working Directory
>
> - Plots / Packages / Help
:::
::::
:::::
:::::::

## Reshuffle #2

- What is your comfort level with R?
- Find a new group of 2-3 people with a mix of R experience
- Introduce yourselves

## Agenda

::: nonincremental
- Math
- Functions
- Data Types
- Logical Operators
- Packages
- Data Structures
- Loading Data
:::

## Math

- At its most basic, R is a calculator
  - *Try it out*: do some math in both the Console and Editor
  - *Shortcut*: [MacOS (Cmd Return) \| Windows (Ctrl Return)]{.smallcaps}

[ ]{.tight-break}

- A **variable** stores a value (use `<-` or `=`)
  - *Try it out*: create some variables and view them in the Environment
  - *Shortcut*: [MacOS (Alt -) \| Windows (Option -)]{.smallcaps}

## Exercise: Math

::: nonincremental
- Pick a number and save it as `x`
- Multiply `x` by 3
- Take the log of the above
- Subtract 4 from the above
- Square the above
:::

## Solution: Math

```{r}
x <- 5
x*3
log(x*3)
(log(x*3))-4
((log(x*3))-4)^2
```

[\* *A quick note on using AI during this program*]{.small-text}

## Functions

- A **function** takes in **parameters** (**arguments**) and outputs values

::: fragment
**Exercise: Functions**
:::

:::: {.fragment .nonincremental .small-text}
::: {}
> Explore the `log` function
>
> - What happens when you type `?log` in the console?
>
> - Take the natural log of 10
>
> - Take the log of 10 with base 2
>
> - Store the results as a variable
:::
::::

## Solution: Functions

```{r}
# Take the natural log of 10
log(10)

# Take the log of 10 with base-2
log(10, base=2)

# Store the result of log(10) as a variable
y <- log(10)
```

## Data Types

- numeric: double, integer
- character: `''` or `""` both work
- logical: `TRUE` vs `FALSE`

[ ]{.tight-break}

- use `class()` or `typeof()` to check on an object's data type

## Data Type Examples

::::: {.columns style="gap: 30px;"}
::: column
```{r}
class(3.14)
class(3)
class(3L)
class("hi")
class(TRUE)
```
:::

::: column
```{r}
typeof(3.14)
typeof(3)
typeof(3L)
typeof("hi")
typeof(TRUE)
```
:::
:::::

## Logical Operators

- These are R's core logical operators

::: {.fragment .nonincremental .small-text}
- `==` for "is equal"
- `!` for "is not"
- `&` for "and"
- `|` for "or"
- `<` for "less than"
- `>` for "greater than"
:::

- A statement with logical operators returns `TRUE` or `FALSE`

## Example: Logical Operators

```{r}
5 > 4

(5 > 4) & (4 < 5)

(5 > 4) | (4 < 5)

!(4 > 5)
```

## Exercise: Logical Operators

::: nonincremental
- Check if 1 is bigger than 2
- Check if 1 + 1 is equal to 2
- Check if the strings "eat" and "drink" are not equal to each other
- Check if it's true that 1 equals 1 *AND* 1 equals 2
- Check if it's true that 1 equals 1 *OR* 1 equals 2
:::

## Solution: Logical Operators

```{r}
1 > 2
(1 + 1) == 2
"eat" != 'drink'
(1==1) & (1==2)
(1==1) | (1==2)
```

## Packages

- **Packages** are additional functions / data sets / tools built by the community, that you can use in R

- Main takeaway: install once and load every time

::: {.fragment .nonincremental}
- **Demo**
  - Install the package `ggplot2`
  - Load the package `ggplot2`
:::

::: fragment
```{r}
#| eval: false
data()
mpg
```
:::

## Basics Review

::: {.fragment .nonincremental}
- Math / Variables
- Functions / Parameters / Arguments
- Data Types
- Logical Operators
- Packages
:::

## BREAK! {.center}

## Reshuffle #3

- What is your work experience?
- Find a new group of 2-3 people with a mix of R experience
- Introduce yourselves

## Data Structures

- How data is structured or organized
- We've been working with a single value so far, but most of the time we'll be working with many values

::: {.fragment .nonincremental}
- Vectors
- Lists
- Factors
- Data Frames
- *Others: Matrices, Arrays, Tibbles, etc.*
:::

## Vectors

- Vectors store multiple values of a **single** data type
- Create a vector with the built-in function `c()`
  - What do I mean by built-in function?
  - What do you think `c` stands for?

::: fragment
```{r}
c(1,2,3,4,5)
vec <- c(1,2,3,4,5)
vec
```
:::

## Vectors

- Vectors can be **indexed** by:
  - element position (`vec[1]`)
  - slice position (`vec[1:3]`)
  - condition (`vec[vec>2]`)

:::: {columns}
::: {.column .fragment style="font-size: 0.8em;"}
```{r}
vec[1]
vec[1:3]
vec[vec>2]
```
:::

::: {.column .small-text .pad-left .fragment}
> How does this compare to other coding languages?
:::
::::

## Example: Vectors

```{r style="font-size: 0.9em;"}
vec <- c("yes", "no", "maybe")
vec

vec <- c(1, 2, 3)
vec

vec[1]

vec*2
log(vec)
```

## Exercise: Vectors

::: nonincremental
- Run the code below to generate `x1` and `x2`
- Select the 3rd element of `x1`
- Select the elements of `x1` less than 0
- Create `x3` with the first five elements of `x2`
- Select all but the fourth element of `x3`

```{r}
set.seed(1234) 
x1 <- rnorm(5)
x2 <- rnorm(20, mean=0.5)
```
:::

## Solution: Vectors

::: nonincremental
```{r}
set.seed(1234) 
x1 <- rnorm(5)
x2 <- rnorm(20, mean=0.5)
x1[3]
x1[x1 < 0]
x3 <- x2[1:5]
x3
x3[-4]
```
:::

## More on Vectors

- When something is **vectorized**, it means that multiple values are processed at once instead of going through a loop

- **Missing values** are stored as `NA` and different functions will handle them differently

::: {.fragment style="font-size: 0.9em;"}
```{r}
vec <- c(1, 8, NA, 7, 3)
mean(vec)
mean(vec, na.rm=TRUE)
sum(is.na(vec)) # the total number of NA values
```
:::

## Lists

- Lists are like vectors, but more flexible
- Lists are *heterogeneous* (vs *homogeneous* vectors): they can store single elements, vectors, or even other lists
- Lists let you keep multi-dimensional or **ragged** data in R
- Index a list element with double brackets: `[[1]]`
  - Single brackets return the element still wrapped in a list

## Example: Lists

```{r}
my_list <- list(42, "Hello world!",  c(1, 2, 3, 4, 5))
my_list[[2]]
my_list[2]
class(my_list[[2]])
class(my_list[2])
```

## Factors

::: {style="font-size: 0.9em;"}
- Factors are a special vector type useful for categorical variables
- Factors have a set number of *levels* (typically auto-detected)
- For categorical variables with a natural order, use *ordered factors*
- Create factors with `factor()`, which takes a `levels` argument
:::

::: {.fragment}
```{r factors}
fac <- factor(c("red", "yellow", "green", "green", "green"))
class(fac)
levels(fac)
```
:::

## Data Frames

- Data frames are R's core data structure: a list of named vectors of equal length
- Data frames are *heterogeneous* — each column can be a different data type
  - Columns are variables
  - Rows are observations
- Build one with `data.frame()` or `cbind()`/`rbind()`
- Extract specific values using `df[row, col]` and specific columns using `df$var` or `df["var"]`

## Example 1: Data Frames
:::fragment
```{r}
df <- data.frame(
  name = c("Ava", "Bob", "Cam"),
  age = c(25, 30, 35),
  active = c(TRUE, FALSE, TRUE)
)
df
department <- c("Sales", "Marketing", "Engineering")
df <- cbind(df, department)
df
```
:::

## Example 1: Data Frames

```{r}
df
df[1, 2]
df$name
df[c("name", "department")]
```

## Example 2: Data Frames

:::fragment
Load an existing Data Frame
```{r}
#| eval: false
data()
data(chickwts)
# click promise

str(chickwts)  # view its structure
head(chickwts) # view the top few rows
View(chickwts) # view the data frame in the UI
```
:::

## Exercise: Data Frames

::: nonincremental
- Load the `mtcars()` data set
- Find the number of rows, columns and column names
- View the data set in the UI
- Select the variable `mpg`
- Select the 4th row
- Square the value of `cyl` and store it as a new variable `cylsq`
:::

## Solution: Data Frames

```{r}
data(mtcars)
str(mtcars)
```

## Solution: Data Frames

```{r style="font-size: 0.9em;"}
mtcars$mpg
mtcars[4,]
mtcars$cylsq <- (mtcars$cyl)^2
head(mtcars)
```

## Reading in Data

- We've covered how to create data frames from scratch and how to load built-in data frames
- In most cases though, we'll want to **load data files** and store them as data frames

- Option 1: Use base R's `read.csv`

::: fragment
```{r style="font-size: 0.9em;"}
df_gm <- read.csv("gapminder.csv")   # ?read.csv for more parameters
head(df_gm)
```
:::

## Reading in Data

- Option 2: Use tidyverse's `readr` package
- In the Files tab, click on the .csv file >> Import Dataset

::: fragment
```{r}
library(readr)
gapminder <- read_csv("gapminder.csv")
head(gapminder, 3)
```
:::

- Notice this returns a tibble, which is a more predictable data frame (more on this later!)

## Discussion

- Combine with another local group
  - What did we cover this morning?
  - Anything stick out in particular?

<br>
  
::: {.fragment}
**CLASS ACTIVITY #1!**
:::

```{=html}
<style>
.small-text {
  font-size: 0.75em;
}

.tight-break {
  line-height: 0.5;
}

.reveal pre {
  margin-top: 10px;
  margin-bottom: 10px;
}

.reveal .column.pad-left {
  padding-left: 50px;
  box-sizing: border-box;
}
</style>
```
