---
title: "Poverty and Test Performance in New York Public Schools"
author: "Puhua Ye"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
source("scripts/3_analyze.R")
source("scripts/4_plot.R")
```

## Question

This project studies the relationship between poverty and test performance in
New York public schools.

## Data and method

The project uses school performance data and county ACS data. Duplicate rows
were removed, and `-99` values were changed to missing values. Counties were
divided into low, medium, and high poverty groups. Test scores were standardized
within each year. The two datasets were joined by county and year.

## Poverty and test performance

```{r poverty-table}
knitr::kable(poverty_group_summary, digits = 2)
```

```{r poverty-plot, fig.width=8, fig.height=5}
poverty_plot
```

Schools in high-poverty counties have lower average reading and math scores.
Schools in low-poverty counties have the highest average scores.

## Change over time

```{r year-plot, fig.width=9, fig.height=5}
year_plot
```

The score gap between low- and high-poverty groups appears in every year. The
size of the gap changes, but it does not disappear.

## Free and reduced-price lunch

```{r lunch-table}
knitr::kable(poverty_lunch_summary, digits = 2)
```

Within each poverty group, schools with higher lunch eligibility have lower
average scores. Lunch eligibility provides additional information about school
poverty and test performance.

## Conclusion

Higher poverty is associated with lower test performance. This relationship
continues over time and is also visible in school lunch eligibility. These
results describe an association and do not prove that poverty causes lower test
scores.
