---
title: "Report"
output: html_document
---

---
title: "New York Schools and County Socioeconomic Conditions"
author: "Yifei Bu"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```

# Introduction

This analysis explores how socioeconomic conditions are related to
academic performance in New York State schools. I focus on county
poverty, student participation in free or reduced-price lunch programs,
and standardized Math and ELA performance.
# Data and Cleaning

The analysis uses school-level data from the New York State Department
of Education and county-level data from the American Community Survey.

Missing values originally coded as -99 were converted to NA. Invalid
percentage values were also treated as missing. Because New York State
test scales changed over time, Math and ELA scores were standardized
within each year using z-scores.

# Poverty Groups

Counties were divided into three approximately equal-sized groups within
each year based on their poverty rate. Counties in the lowest third were
classified as low poverty, those in the middle third as medium poverty,
and those in the highest third as high poverty.

```{r load-results}
county_year_summary <- read_csv(
  "data/processed/county_year_summary.csv",
  show_col_types = FALSE
)

poverty_extremes <- read_csv(
  "data/processed/poverty_extremes.csv",
  show_col_types = FALSE
)

poverty_group_summary <- read_csv(
  "data/processed/poverty_group_summary.csv",
  show_col_types = FALSE
)
```
# Summary Tables

## Counties with the Highest and Lowest Poverty Rates
```{r poverty-table}
knitr::kable(
  poverty_extremes,
  digits = 2,
  caption = "Counties with the Highest and Lowest Average Poverty Rates"
)
```

```{r create-plots, include=FALSE}
source("scripts/4_plot.R")
```
# Lunch Assistance and Academic Performance
```{r plot1, echo=FALSE, fig.width=10, fig.height=5}
p1
```


# Key Findings

The analysis shows clear socioeconomic differences in school outcomes.
Schools serving larger shares of students receiving free or reduced-price
lunch tend to have lower standardized test performance. Academic
performance also differs across county poverty groups.

These results describe associations in the data and should not be
interpreted as evidence that poverty or lunch-program participation
directly causes differences in test scores.
