---
title: "Exploring the Correlation between poverty and standardized test grades"
output: html_document
date: "2026-09-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, error = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
data <- read_csv('data/processed/schools.csv')
```

# County's basic information 


```{r}
summary_1 <- data |>
  group_by(county_name)|>
  summarize(total_enrollment = sum(total_enroll),
            mean_poverty = mean(county_per_poverty))
summary_1
```


# Comparing test scores across different poverty level

Because the scoring system varies across each years, tests score each years are standardized into z-scores based on scores from all schools that year so that they could be compared across multiple years.

## Standardized Math Scores
```{r}
data |>
  mutate(poverty_level = factor(poverty_level,
                                   levels = c('Low','Medium','High'))) |>
  ggplot(aes(x = poverty_level, y = std_math)) +
  geom_boxplot()+
  theme_minimal()+
  labs(title = 'High poverty groups tend to have lower median math score',
       subtitle = 'Poverty level defined by percent of poverty of each county in each year',
       x = 'poverty level',
       y = 'standardized math tests score')
```

## Standardized English Language Scores
```{r}
data |>
  mutate(poverty_level = factor(poverty_level,
                                levels = c('Low','Medium','High'))) |>
  ggplot(aes(x = poverty_level, y = std_ela)) +
  geom_boxplot()+
  theme_minimal()+
  labs(title = 'High poverty groups tend to have lower median score',
       subtitle = 'Poverty level defined by percent of poverty of each county in each year',
       x = 'poverty level',
       y = 'standardized reading tests score')
```

# Correlation between 2 test scores

```{r}
data |>
  mutate(poverty_level = factor(poverty_level,
                                   levels = c('Low','Medium','High'))) |>
  ggplot(aes(x = std_ela, y = std_math, color = poverty_level))+
  geom_point(alpha = 0.2)+
  theme_minimal()+
    labs(title = 'There is a strong correlations between two test scores',
       subtitle = 'The upper tail with both high scores is mostly of high poverty schools',
       x = 'standardized reading tests score',
       y = 'standardized math tests score')
```





